### 2016-05-21 papers/thesis-vischer/Thesis.tex
add note about Stirling

We begin by imagining a cell of length $L$ and volume $L^3$ in which lives a particle. Then, we quantize the allowed wavelengths of the particle within the cell by defining the wavevector $\mathbf{k}$ as
\begin{align}
    \mathbf{k} &= \frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}})
\end{align}
Remembering that kinetic energy can be expressed as $K = \frac{\hbar^2 k^2}{2m}$, and using our definition of $\mathbf{k}$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_s V\, \diff\mathbf{p} ~e^{-\beta \frac{\hbar^2 k^2}{2m}}\right]^{N}
    \label{Zt_def}
\end{align}
This unfortunately introduces a mass quantity into the partition function which adds an additional parameter that needs to be determined to properly account for quantum mechanical effects. If we then allow the cell to become very large, then the we can easily see in Eq. \ref{Zt_def} that the spacing between occupants in phase space becomes small, so we can integrate instead $\diff ^3 n$ over all possible values of $n$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{-\frac{\beta \hbar^2}{2m} \left(\frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}}\right)^2}\right]^{N} \\
    &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{ \frac{-2\beta \hbar^2\pi^2}{mL^2}(n_x^2 + n_y^2 + n_z^2)}\right]^{N}
\end{align}
where extra factors of $\hbar $and$ \pi$ were gained from the change of variables. Evaluating this integral yields
\begin{align}
    Z_{id} &= \left(\frac{V^N}{N!\Lambda^{3N}}\right) \,, \\
    \Lambda &= \frac{\hbar\sqrt{2\pi}}{\sqrt{mT}}
\end{align}
where we have introduced the thermal deBroglie wavelength $\Lambda$ which defines the de Broglie wavelength of an ideal gas at a temperature $T$. The full partition function for any monoatomic system is then
\begin{align}
    Z &= Z_{exc} \left(\frac{V}{N!\Lambda^{3}}\right)^{N} \,.
\end{align}
For liquids, the $Z_{exc}$ factor is difficult to evaluate; the hard sphere liquid, the simplest model for a dense fluid, does not have an analytic representation for the excess free energy. It follows that a considerable amount of work is required to determine the excess free energy for a given system. The absolute free energy, by way of Eq. \ref{FdefSM}, now has the form
\begin{align}
    F_{abs} &= -T\,\ln\left(Z_{exc} \frac{V^N}{N!\Lambda^{3N}}\right)\\
    &= -T\ln\left(Z_{exc}\right)\, -T\left(\frac{V}{(N!)^{1/N}\Lambda^3} \right)^N\\
    &= -T\ln\left(Z_{exc}\right) - NT\ln\left(\frac{V}{\Lambda^3}\right) + NT\ln\left(N!^{1/N}\right)\\ 
\end{align}
Redefining the first factor including $Z_{exc}$ to be $F_{exc}$, and then using Sterling's approximation ($\ln(N!) = N\ln N$ for large N), we arrive at
\begin{align}
    F_{abs} &= F_{exc} - NT\left(\ln\left(\frac{V}{\Lambda^3}\right) - \ln\left(N\right)\right)
\end{align}
At a specified volume and number of particles, the ideal gas contribution is only a function of temperature. Consequently, for a given temperature we only need to compute the excess free energy.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                            METHODS                                  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

We begin by imagining a cell of length $L$ and volume $L^3$ in which lives a particle. Then, we quantize the allowed wavelengths of the particle within the cell by defining the wavevector $\mathbf{k}$ as
\begin{align}
    \mathbf{k} &= \frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}})
\end{align}
Remembering that kinetic energy can be expressed as $K = \frac{\hbar^2 k^2}{2m}$, and using our definition of $\mathbf{k}$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_s V\, \diff\mathbf{p} ~e^{-\beta \frac{\hbar^2 k^2}{2m}}\right]^{N}
    \label{Zt_def}
\end{align}
This unfortunately introduces a mass quantity into the partition function which adds an additional parameter that needs to be determined to properly account for quantum mechanical effects. If we then allow the cell to become very large, then the we can easily see in Eq. \ref{Zt_def} that the spacing between occupants in phase space becomes small, so we can integrate instead $\diff ^3 n$ over all possible values of $n$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{-\frac{\beta \hbar^2}{2m} \left(\frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}}\right)^2}\right]^{N} \\
    &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{ \frac{-2\beta \hbar^2\pi^2}{mL^2}(n_x^2 + n_y^2 + n_z^2)}\right]^{N}
\end{align}
where extra factors of $\hbar $and$ \pi$ were gained from the change of variables. Evaluating this integral yields
\begin{align}
    Z_{id} &= \left(\frac{V^N}{N!\Lambda^{3N}}\right) \,, \\
    \Lambda &= \frac{\hbar\sqrt{2\pi}}{\sqrt{mT}}
\end{align}
where we have introduced the thermal deBroglie wavelength $\Lambda$ which defines the de Broglie wavelength of an ideal gas at a temperature $T$. The full partition function for any monoatomic system is then
\begin{align}
    Z &= Z_{exc} \left(\frac{V}{N!\Lambda^{3}}\right)^{N} \,.
\end{align}
For liquids, the $Z_{exc}$ factor is difficult to evaluate; the hard sphere liquid, the simplest model for a dense fluid, does not have an analytic representation for the excess free energy. It follows that a considerable amount of work is required to determine the excess free energy for a given system. The absolute free energy, by way of Eq. \ref{FdefSM}, now has the form
\begin{align}
    F_{abs} &= -T\,\ln\left(Z_{exc} \frac{V^N}{N!\Lambda^{3N}}\right)\\
    &= -T\ln\left(Z_{exc}\right)\, -T\left(\frac{V}{(N!)^{1/N}\Lambda^3} \right)^N\\
    &= -T\ln\left(Z_{exc}\right) - NT\ln\left(\frac{V}{\Lambda^3}\right) + NT\ln\left(N!^{1/N}\right)\\ 
\end{align}
Redefining the first factor including $Z_{exc}$ to be $F_{exc}$, and then using Sterling's approximation ($\ln(N!) = N\ln N$ for large N), we arrive at \textcolor{red}{BRENDEN FIX: $N\ln N - N$, plus I think we may want to add to the excess free energy (probably to the excess HS entropy?) the difference?}
\begin{align}
    F_{abs} &= F_{exc} - NT\left(\ln\left(\frac{V}{\Lambda^3}\right) - \ln\left(N\right)\right)
\end{align}
At a specified volume and number of particles, the ideal gas contribution is only a function of temperature. Consequently, for a given temperature we only need to compute the excess free energy.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                            METHODS                                  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
